Parallelizing Linear Recurrent Neural Nets Over Sequence Length

### 2017-10-27 paper/main.tex
Cleaned up plots, added extra diagram

\documentclass{article}
\usepackage{iclr2018_conference,times}
\bibliographystyle{abbrvnat}

@@

\begin{abstract}
Recurrent neural networks (RNNs) are widely used to model sequential data but
their non-linear dependencies between sequence elements prevent parallelizing
training over sequence length. We show the training of RNNs with only linear
sequential dependencies can be parallelized over the sequence length using the
parallel scan algorithm, leading to rapid training on long sequences even with small
minibatch size. We develop a parallel linear recurrence CUDA kernel and show that
it can be applied to immediately speed up training and inference of
several RNN architectures by up to 9x.
We abstract recent work on linear RNNs into a new framework of linear surrogate RNNs
and develop
a linear surrogate model for the long short-term memory unit, the GILR-LSTM, that
utilizies parallel linear recurrence.
We push the limits of sequence learning by successfully training a GILR-LSTM
on a synthetic sequence classification task with a one million timestep dependency.
SOMETHING MEDICAL?
\end{abstract}

\section{Introduction}
Recurrent neural networks (RNNs) are widely used for sequence modelling tasks in
domains such as natural language processing \cite{sutskever2014sequence}, speech
recognition \cite{amodei2015deep}, and reinforcement learning
\cite{hausknecht2015deep}. Most RNNs, including popular variants such as long
short-term memories (LSTMs) \cite{hochreiter1997long} and gated recurrent units
(GRUs) \cite{cho2014learning}, contain a non-linear dependency between
sequential inputs. These non-linear dependencies create a very flexible class of
models but limit the feasibility of training RNNs on long sequences as each
sequence element must be processed sequentially.  Modelling sequences of
thousands to millions of elements is important to domains such as robotics,
remote sensing, control systems, speech recognition, medicine, and finance.

The RNN serial evaluation inefficiency problem is usually mitigated by
parallelizing the forward and backward pass over a minibatch of inputs. Without
minibatches, RNN evaluation is a sequence of matrix-vector
multiplications. Minibatches transform RNN computation into a sequence of more
efficient matrix-matrix multiplications, but this speed-up brings several
disadvantages.  RNN model size is often limited by GPU memory size, and running
a forward and backward pass on a minibatch requires memory linear in the
minibatch size.  Grouping data into minibatches increases the latency of each
pass and reduces the rate of optimization steps. Finally, training with larger
minibatches damages generalization ability \cite{keskar2017large}. Given these
effects, it is desirable to obtain high training throughput with small minibatches.
Persistent RNNs \cite{diamos2016persistent} use a novel
implementation that can achieve high GPU utilization with very small minibatch
sizes when the recurrent state is larger than 500 elements, but even persistent
RNNs become limited by the serial evaluation inefficiency at smaller hidden
sizes.

@@

A linear recurrence is a specific instance of a general form of computation known as a scan.
Scans and reductions are computations involving repeated application of a binary
operator $\oplus$ over an array of data. Computing the sum or maximum
of an array is an example of a reduction, while a cumulative sum is a common
example of a scan operation. Throughout this work, the scan of $\oplus$ with
initial value $b$ is defined as
\begin{align*}
\scan(\oplus, [a_1, a_2, ..., a_n], b) = [(a_1 \oplus b), (a_2 \oplus a_1 \oplus b), ..., (a_n \oplus a_{n-1} ... \oplus a_1 \oplus b)]
\end{align*}
The reduction of $\oplus$ over array $A$ and initial value $b$ is denoted
$\reduce(\oplus, A, b)$ and is the final element of $\scan(\oplus, A, b)$.
Despite their dependent computation graph, algorithms exist to parallelize scans
and reductions when $\oplus$ is associative \cite{ladner1980parallel}.

@@

We investigate this idea further and deliver the following contributions:
\begin{itemize}
\item{We classify RNNs which satisfy the conditions above, and
    show that many RNNs used in practice such as Quasi-RNNs (QRNNs) \cite{bradbury2017quasi} are
    contained in this class}
\item{We provide an implementation of the parallel linear recurrence
    algorithm as a CUDA kernel, and show that it speeds up training of
    QRNN and Simple Recurrent Unit (SRU) \cite{lei2017} architectures by factors of up to 9x}
\item{We describe how several recent linear RNNs can be described as
    linear surrogates for non-linear architectures. We introduce a linear
    surrogate for the LSTM and show that we are able to train it
    with a speedup of 5-10x compared to the CuDNN-LSTM when we use the
    parallel linear recurrence algorithm.}
\end{itemize}

@@

\subsection{Theoretical performance}
The cost of a serial scan over a sequence of length $T$ is
$C_\text{sscan} \in \mathcal{O}((C_\otimes + C_\oplus)T)$, compared to the parallel scan cost
$C_\text{pscan} \in \mathcal{O}(2(C_\odot + C_\otimes + C_\oplus)(T/p + \lg p))$ \cite{blelloch1990prefix} on $p$ processors.
If $h_t$ is a vector of dimension $n$ then
$C_\odot \in  \mathcal{O}(n^3), C_\otimes \in \mathcal{O}(n^2), C_\oplus \in \mathcal{O}(n)$ giving
$C_\text{pscan} \in \mathcal{O}(2(n^3 + n^2 + n)(T/p + \lg p))$ and
$C_\text{sscan} \in \mathcal{O}((n^2 + n)T)$. The $\mathcal{O}(n^3)$ cost of the matrix
multiplication in the parallel algorithm can counter-act any parallel speedups for
sufficiently large hidden states and lead to a slower algorithm overall.

To avoid this problem, we will only consider diagonal matrices $\Lambda_t$, in
which case both matrix-matrix and matrix-vector multiplication have cost
proportional to $n$ and $C_\text{pscan}\in \mathcal{O}(6n(T/p + \lg p))$ and
$C_\text{sscan} \in \mathcal{O}(2nT)$. This gives a parallel speedup factor of
$\frac{pT}{3(T+\lg p)}$. Assuming $p \ll T$, then $C_\text{pscan} \le
C_\text{sscan}$ when $p \ge 3$.

@@

Limiting $\Lambda_t$ to be diagonal may seem like a severe constraint but there are
several reasons to do so beyond the unfavorable parallelization performance. Relatively few neural
network models use separate recurrent matrices for each sequence element and using these
separate matrices would require potentially prohibitive $n^2T$ memory.
Applying
the same matrix $\Lambda$ to each sequence element is also unappealing considering that a matrix
multiplication can be thought of as a rotation and a scaling. The same rotation at every
element seems unlikely to be useful, and the scaling is exactly what's captured in diagonal
vectors $\lambda_t$. Recurrent coefficient vectors $\lambda_t$ provide enough flexibility
to implement schemes such as exponential moving averages or a gating mechanism.

@@

This work implemented parallel linear recurrence as a CUDA kernel with bindings
into the TensorFlow \cite{abadi2016tensorflow} framework. Each warp acts as a
processor, which means the algorithmic $p$ is up to 3200 and the theoretical
parallelization speedup factor is up to several hundred.  The 32 lanes of each
warp work on different elements of the recurrence vector in parallel. These
implementation details mean that peak performance is only obtained on sequences
of at least several thousand steps on at least a 32 element vector.

@@

\subsection{Linear surrogate RNNs}
\label{sec:ls-rnns}
RNNs learn a transition function $s_t = f(s_{t-1}, x_t)$ which combines previous
state $s_{t-1}$ with input $x_t$ to compute current state $s_t$. Non-linear $f$
prevents application of the parallel linear recurrence algorithm and forces slow
serial evaluation. To work around this inefficiency, note that $s_t$ serves
dual purposes. In $s_t = f(s_{t-1}, x_t)$, $s_{t-1}$ serves as an input to $f$
summarizing the previous inputs while $s_t$ serves as the output of $f$ to be
passed to other layers of the network. We can decouple these uses and introduce
independent variables for each purpose: \(s_t\) is passed onto other layers of the network
and we introduce the linear surrogate \(\tilde{s}_t\) which is passed onto the
next state, with \(s_t = f(\tilde{s}_{t-1}, x_t)\). We are still able to choose a
non-linear \(f\), our only limitation being that \(\tilde{s}_t\) must be linearly
computable.  We refer to this class of model as a linear surrogate RNN
(LS-RNN). QRNNs \cite{bradbury2017quasi} are LS-RNNs using $\tilde{h}_{t-1}
= W_k x_{t-k} + ... W_1 x_{t-1}$ and strongly typed
RNNs\cite{balduzzi2016strongly} are LS-RNNs with $\tilde{h}_t=x_{t-1}$. Although
not a rule, LS-RNNs can often be parallelized over sequence length with either
convolution or linear recurrence.

Consider an LSTM:
\begin{align*}
  f_t, i_t, o_t &= \sigma(U_{f,i,o} h_{t-1} + V_{f,i,o} x_t + b_{f,i,o}) \\
  z_t &= \tau(U_z h_{t-1} + V_z x_t + b_z) \\
  c_t &= f_t \odot c_{t-1} + i_t \odot z_t \\
  h_t &= o_t \odot c_t
\end{align*}
An LSTM has state $s_t = (h_t, c_t)$. Since $c_t$ depends only
linearly on $c_{t-1}$, no surrogate is needed for $c_t$. $h_t$ has a non-linear
dependence on $h_{t-1}$, so $h_t$ needs a linear surrogate. Introducing a GILR layer as
the surrogate, we obtain the GILR-LSTM:
\begin{align*}
  g_t &= \sigma(V_g x_t + b_g) \\
  j_t &= \tau(V_j x_t + b_j) \\
  \tilde{h}_t &= g_t \odot \tilde{h}_{t-1} + (1-g_t)\odot j_t \\
  f_t, i_t, o_t &= \sigma(U_{f,i,o} \tilde{h}_{t-1} + V_{f,i,o} x_t + b_{f,i,o}) \\
  z_t &= \tau(U_z \tilde{h}_{t-1} + V_z x_t + b_z) \\
  c_t &= f_t \odot c_{t-1} + i_t \odot z_t \\
  h_t &= o_t \odot c_t \\
\end{align*}

@@

In order to illustrate that the linearization does not necessarily come at the
cost of expressibility, we show that the GILR-LSTM architecture computed with
the parallel linear recurrence algorithm is able to train significantly faster
than an optimized LSTM implementation on a pathological long-term dependency
problem on a pathological example from the original LSTM paper \cite{hochreiter1997long}.
Finally we show that the GILR-LSTM is able to
quickly achieve state of the art results on an existing medical data
set. \Chris{I hope so\ldots}

@@

\subsubsection{Accelerating existing RNN architectures}
Several recently introduced LS-RNNs can be accelerated with the parallel linear recurrence
algorithm. We implemented SRUs, QRNNs (with filter width 2 and 10), and GILR-LSTMs
that can be computed with either the standard
serial linear recurrence algorithm or parallel linear recurrence. Both methods compute
an identical recurrence, so switching from a serial to parallel implementation
does not cause any numerical changes and takes only a single line of code changes.
Notably, both SRUs and QRNNs
claim an order of magnitude speedup compared to CuDNN LSTM when implemented with serial linear
recurrence. Any further speedup from parallel linear recurrence applies on top of the existing
speedup.
We timed train throughput (forwards and backwards propagation), but the linear time
of each pass also makes the results applicable to forwards (inference) performance. However,
parallel linear recurrence can only accelerate inference in scenarios where the entire
input sequence is known at the start of the inference phase.
We controlled for GPU memory usage within these experiments by fixing $bT = 65536$ for
minibatch size $b$ and sequence length $T$.

Table \ref{table:rnn-throughput} shows that
throughput advantage from using parallel linear recurrence compared to
serial linear recurrence reaches up to 9x. Simpler architectures
(for which the linear recurrence is a higher proportion of the
total computational load) are more affected by the switch to the parallel kernel.
This is
particularly clear in the case of the QRNN, where including wider
convolutional filters results in more time spent outside of the linear recurrence and
therefore speedup from linear recurrence parallelization.
\begin{table}[t]
\caption{Parallel kernel speedup for a variety of LS-RNNs}
\begin{center}
\begin{tabular}{@{}lrrrr@{}}
\label{table:rnn-throughput}
Sequence Length & SRU & QRNN (filter size 2) & QRNN (filter size 10) & GILR-LSTM\\ \midrule
16 & 0.28 & 0.38 & 0.78 & 0.61\\
256 & 0.84 & 0.86 & 0.99 & 0.91\\
4,096 & 1.38 & 1.18 & 1.05 & 0.98\\
65,536 & 9.21 & 6.68 & 2.05 & 1.41\\ \bottomrule
\end{tabular}
\end{center}
\end{table}

\Eric{For table \ref{table:rnn-throughput} how many layers and what was hidden size?}

\subsubsection{GILR-LSTM throughput}

\Eric{I think we should delete this section. Already over length limit, doesn't show
anything great}

Eric: I think we should delete this. We're over length, results aren't great.

Finally, we compare the throughput of the GILR-LSTM to the current fastest
LSTM implementation (which comes from the CuDNN library). Figure
\ref{fig:tp_perf} shows that the GILR-LSTM has higher throughput for most of the
small minibatch regime, with a 30x speedup at a sequence length of 65,536.
The GILR-LSTM allows us to achieve high throughput for all choices of minibatch size and
sequence length.

Unlike the previous case, these speedups do come at a cost. The GILR-LSTM
architecture is different to the canonical LSTM, and it's possible that the loss
of model capacity outweighs the speedup in training. The following experiment
shows that the GILR-LSTM is able to outperform the LSTM on an example originally
designed to illustrate the advantages of the LSTM.

\Eric{Chris: Put title on Figure \ref{fig:tp_perf}}

\begin{figure}[t]
\centering
\includegraphics[width=12cm]{cudnn_heatmap.png}
\caption{Throughput comparison between LSTM-256-256 with GILR-LSTM-256-256, with a
  32-dimensional input and a 2-dimensional output, for various batch sizes and
  sequence lengths.The LSTM only has a single row of data because its throughput
  is independent of sequence length. Entries are missing from the GILR-LSTM table
  because there was not enough memory on the GPU to handle such large batch
  sizes and sequences.}
\label{fig:tp_perf}
\end{figure}

\subsection{Synthetic Experiment}
In order to demonstrate how PLR may be used to speed up tasks that the LSTM is
well-suited for, we tackle a pathological inference problem first introduced
in Hochreiter and Schmidhuber's initial LSTM paper \cite{hochreiter1997long}.
The input consists of sequences of length $n$ with each sequence element a randomly
chosen one-hot vector in $p$ dimensional space.
The first vector in each sequence is always the same, up to the sign of the component (i.e. it is $\pm
\boldsymbol{p_0}$). The sequential model must read in an entire sequence and then output
the sign of the first sequence element.
This sequence classification problem requires remembering the first element over the length of the
sequence and is a canonical example of the long-term dependency problem.
Due to the large throughput of the GILR-LSTM, we would expect that it
would do better when the sequence length is large. In the original formulation
of the problem (dealing in the regime with around one hundred timesteps)
, $p$ is set equal to $n$. As this would make the size of the input data
grow impracticaly large as \(\mathcal{O}(n^2)\) for long sequences, we fix \(p =
128\) and vary \(n\).

We generated sequences for $n$ equal to 1,024, 8,192, and
1,048,576. For each of these we compared a two layer GILR-LSTM with 512 hidden
units to a two layer LSTM with 512 hidden units per layer implemented by CuDNN.

We ran all experiments on a NVIDIA K80 GPU, choosing the largest
minibatch size which fit into the GPU memory with five runs per configuration
allowing us to find the average and standard deviation of the time and number of
iterations to convergence. A brief search over learning rate
and batch size was carried out to find the parameters which allow the network to
converge most rapidly for all runs. The criterion for convergence was five consecutive
minibatches giving 100\% accuracy. The learning curves
in figure \ref{fig:synthetic_training} give support to this being a reasonable
convergence criteria.
For the longest sequence length, we could not run long enough to observe the LSTM converge.

The results illustrate that the GILR-LSTM is able to converge much more quickly than
the fastest available LSTM implementation.
This is somewhat surprising given the LSTM was specifically constructed for problems
of this sort. This GILR-LSTM convergence
leads to the conclusion that the non-linearities present in
LSTM are not necessary for solving this instance of the long-term dependency problem.
The time to convergence further leads to the conclusion that inclusion of a non-linearity
at every step incurs a significant training time slowdown.
Furthermore, the GILR-LSTM is able to learn to carry dependencies over a one million
element sequence. As far as we know, this one million step sequence experiment
is the longest sequential learning problem to be handled by neural networks to date.

\begin{figure}
  \centering
  \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./1k_synthetic.png}
  \end{minipage}%
    \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./8k_synthetic.png}
  \end{minipage}
  \begin{minipage}{0.6\textwidth}
    \includegraphics[width=1.0\textwidth]{./1m_synthetic.png}
  \end{minipage}
  \caption{Training curves for GILR-LSTM and CUDNNLSTM architectures for various
  sequence lengths. (Clockwise from top-left: 1024, 8192, 1048576)}
    \label{fig:synthetic_training}
  \end{figure}

>>>>

\documentclass{article}
\usepackage{iclr2018_conference,times}
\bibliographystyle{natbib}

@@

\begin{abstract}
Recurrent neural networks (RNNs) are widely used to model sequential data but
their non-linear dependencies between sequence elements prevent parallelizing
training over sequence length. We show the training of RNNs with only linear
sequential dependencies can be parallelized over the sequence length using the
parallel scan algorithm, leading to rapid training on long sequences even with
small minibatch size. We develop a parallel linear recurrence CUDA kernel and
show that it can be applied to immediately speed up training and inference of
several state of the art RNN architectures by up to 9x.  We abstract recent work
on linear RNNs into a new framework of linear surrogate RNNs and develop a
linear surrogate model for the long short-term memory unit, the GILR-LSTM, that
utilizies parallel linear recurrence.  We extend sequence learning to new
extremely-long-timestep regimes that were previously out of reach by
successfully training a GILR-LSTM on a synthetic sequence classification task
with a one million timestep dependency.
\end{abstract}

\section{Introduction}
Recurrent neural networks (RNNs) are widely used for sequence modelling tasks in
domains such as natural language processing \citep{sutskever2014sequence},
speech recognition \citep{amodei2015deep}, and reinforcement learning
\citep{hausknecht2015deep}. Most RNNs, including popular variants such as long
short-term memories (LSTMs), introduced by \citet{hochreiter1997long}, and gated
recurrent units (GRUs), introduced by \citet{cho2014learning}, contain a non-linear dependency
between sequential inputs. These non-linear dependencies create a very flexible
class of models but limit the feasibility of training RNNs on long sequences as
each sequence element must be processed sequentially.  Modelling sequences of
thousands to millions of elements is important to domains such as robotics,
remote sensing, control systems, speech recognition, medicine, and finance.

The RNN serial evaluation inefficiency problem is usually mitigated by
parallelizing the forward and backward pass over a minibatch of inputs. Without
minibatches, RNN evaluation is a sequence of matrix-vector
multiplications. Minibatches transform RNN computation into a sequence of more
efficient matrix-matrix multiplications, but this speed-up brings several
disadvantages.  RNN model size is often limited by GPU memory size, and running
a forward and backward pass on a minibatch requires memory linear in the
minibatch size.  Grouping data into minibatches increases the latency of each
pass and reduces the rate of optimization steps. Finally, training with larger
minibatches damages generalization ability \citep{keskar2017large}. Given these
effects, it is desirable to obtain high training throughput with small minibatches.
Persistent RNNs \citep{diamos2016persistent} use a novel
implementation that can achieve high GPU utilization with very small minibatch
sizes when the recurrent state is larger than 500 elements, but even persistent
RNNs become limited by the serial evaluation inefficiency at smaller hidden
sizes.

@@

A linear recurrence is a specific instance of a general form of computation known as a scan.
Scans and reductions are computations involving repeated application of a binary
operator $\oplus$ over an array of data. Computing the sum or maximum
of an array is an example of a reduction, while a cumulative sum is a common
example of a scan operation. Throughout this work, the scan of $\oplus$ with
initial value $b$ is defined as
\begin{align*}
\scan(\oplus, [a_1, a_2, ..., a_n], b) = [(a_1 \oplus b), (a_2 \oplus a_1 \oplus b), ..., (a_n \oplus a_{n-1} ... \oplus a_1 \oplus b)].
\end{align*}
The reduction of $\oplus$ over array $A$ and initial value $b$ is denoted
$\reduce(\oplus, A, b)$ and is the final element of $\scan(\oplus, A, b)$.
Despite their dependent computation graph, algorithms exist to parallelize scans
and reductions when $\oplus$ is associative \citep{ladner1980parallel}.

@@

We investigate this idea further and deliver the following contributions:
\begin{itemize}
\item{We classify RNNs which satisfy the conditions above, and show that many
    RNNs used in practice such as the Quasi-RNNs (QRNNs) introduced by
    \citet{bradbury2017quasi} are contained in this class.}
\item{We provide an implementation of the parallel linear recurrence algorithm
    as a CUDA kernel, and show that it speeds up training of QRNN and
    \citet{lei2017}'s Simple Recurrent Unit (SRU) architectures by factors of up
    to 9x.}
\item{We describe how several recent linear RNNs can be described as
    linear surrogates for non-linear architectures. We introduce a linear
    surrogate for the LSTM and show that we are able to train it
    with a speedup of 5-10x compared to the CuDNN-LSTM when we use the
    parallel linear recurrence algorithm.}
\end{itemize}

@@

\subsection{Theoretical performance}
The cost of a serial scan over a sequence of length $T$ is
$C_\text{sscan} \in \mathcal{O}((C_\otimes + C_\oplus)T)$, compared to the parallel scan cost
$C_\text{pscan} \in \mathcal{O}(2(C_\odot + C_\otimes + C_\oplus)(T/p + \lg p))$ on $p$ processors \citep{blelloch1990prefix}.
If $h_t$ is a vector of dimension $n$ then
$C_\odot \in  \mathcal{O}(n^3), C_\otimes \in \mathcal{O}(n^2), C_\oplus \in \mathcal{O}(n)$ giving
$C_\text{pscan} \in \mathcal{O}(2(n^3 + n^2 + n)(T/p + \lg p))$ and
$C_\text{sscan} \in \mathcal{O}((n^2 + n)T)$. The $\mathcal{O}(n^3)$ cost of the matrix
multiplication in the parallel algorithm can counter-act any parallel speedups for
sufficiently large hidden states and lead to a slower algorithm overall.

To avoid this problem, we will only consider diagonal matrices $\Lambda_t$, in
which case both matrix-matrix and matrix-vector multiplication have cost
proportional to $n$ and $C_\text{pscan}\in \mathcal{O}(6n(T/p + \lg p))$ and
$C_\text{sscan} \in \mathcal{O}(2nT)$. This gives a parallel speedup factor of
$pT/3(T+\lg p)$. Assuming $p \ll T$, then $C_\text{pscan} \le
C_\text{sscan}$ when $p \ge 3$.

@@

Limiting $\Lambda_t$ to be diagonal may seem like a severe constraint but there are
several reasons to do so beyond the favorable parallelization performance. Relatively few neural
network models use separate recurrent matrices for each sequence element and using these
separate matrices would require potentially prohibitive $n^2T$ memory.
Applying
the same matrix $\Lambda$ to each sequence element is also unappealing considering that a matrix
multiplication can be thought of as a rotation and a scaling. The same rotation at every
element seems unlikely to be useful, and the scaling is exactly what's captured in diagonal
vectors $\lambda_t$. Recurrent coefficient vectors $\lambda_t$ provide enough flexibility
to implement schemes such as exponential moving averages or a gating mechanism.

@@

This work implemented parallel linear recurrence as a CUDA kernel with bindings
into the TensorFlow \citep{abadi2016tensorflow} framework. Each warp acts as a
processor, which means the algorithmic $p$ is up to 3200 and the theoretical
parallelization speedup factor is up to several hundred.  The 32 lanes of each
warp work on different elements of the recurrence vector in parallel. These
implementation details mean that peak performance is only obtained on sequences
of at least several thousand steps on at least a 32 element vector.

@@

\subsection{Linear surrogate RNNs}
\label{sec:ls-rnns}
RNNs learn a transition function $s_t = f(s_{t-1}, x_t)$ which combines previous
state $s_{t-1}$ with input $x_t$ to compute current state $s_t$. Non-linear $f$
prevents application of the parallel linear recurrence algorithm and forces slow
serial evaluation. To work around this inefficiency, note that $s_t$ serves
dual purposes. In $s_t = f(s_{t-1}, x_t)$, $s_{t-1}$ serves as an input to $f$
summarizing the previous inputs while $s_t$ serves as the output of $f$ to be
passed to other layers of the network. We can decouple these uses and introduce
independent variables for each purpose: \(s_t\) is passed onto other layers of the network
and we introduce the linear surrogate \(\tilde{s}_t\) which is passed onto the
next state, with \(s_t = f(\tilde{s}_{t-1}, x_t)\). We are still able to choose a
non-linear \(f\), our only limitation being that \(\tilde{s}_t\) must be linearly
computable.  We refer to this class of model as a linear surrogate RNN
(LS-RNN). QRNNs \citep{bradbury2017quasi} are LS-RNNs using $\tilde{h}_{t-1}
= W_k x_{t-k} + ... W_1 x_{t-1}$ and strongly typed
RNNs \citep{balduzzi2016strongly} are LS-RNNs with $\tilde{h}_t=x_{t-1}$. Although
not a rule, LS-RNNs can often be parallelized over sequence length with either
convolution or linear recurrence.

Consider an LSTM:
\begin{align*}
  f_t, i_t, o_t &= \sigma(U_{f,i,o} h_{t-1} + V_{f,i,o} x_t + b_{f,i,o}) \\
  z_t &= \tau(U_z h_{t-1} + V_z x_t + b_z) \\
  c_t &= f_t \odot c_{t-1} + i_t \odot z_t \\
  h_t &= o_t \odot c_t
\end{align*}
An LSTM has state $s_t = (h_t, c_t)$. Since $c_t$ depends only
linearly on $c_{t-1}$, no surrogate is needed for $c_t$. $h_t$ has a non-linear
dependence on $h_{t-1}$, so $h_t$ needs a linear surrogate. Introducing a GILR layer as
the surrogate, we obtain the GILR-LSTM:

\begin{align*}
  g_t &= \sigma(V_g x_t + b_g) \\
  j_t &= \tau(V_j x_t + b_j) \\
  \tilde{h}_t &= g_t \odot \tilde{h}_{t-1} + (1-g_t)\odot j_t \\
  f_t, i_t, o_t &= \sigma(U_{f,i,o} \tilde{h}_{t-1} + V_{f,i,o} x_t + b_{f,i,o}) \\
  z_t &= \tau(U_z \tilde{h}_{t-1} + V_z x_t + b_z) \\
  c_t &= f_t \odot c_{t-1} + i_t \odot z_t \\
  h_t &= o_t \odot c_t \\
\end{align*}

@@

In order to illustrate that the linearization does not necessarily come at the
cost of expressibility, we show that the GILR-LSTM architecture computed with
the parallel linear recurrence algorithm is able to train significantly faster
than an optimized LSTM implementation on a pathological long-term dependency
problem from the original LSTM paper \citep{hochreiter1997long}.
Finally we show that the GILR-LSTM is able to
quickly achieve state of the art results on an existing medical data
set. \Chris{I hope so\ldots}

@@

\subsubsection{Accelerating existing RNN architectures}
Several recently introduced LS-RNNs can be accelerated with the parallel linear
recurrence algorithm. We implemented SRUs, QRNNs (with filter width 2 and 10),
and GILR-LSTMs that can be computed with either the standard serial linear
recurrence algorithm or parallel linear recurrence. Both methods compute an
identical recurrence, so switching from a serial to parallel implementation does
not cause any numerical changes and takes only a single line of code changes.
Notably, both SRUs and QRNNs claim an order of magnitude speedup compared to
CuDNN LSTM when implemented with serial linear recurrence. Any further speedup
from parallel linear recurrence applies on top of the existing speedup.  We
timed train throughput (forwards and backwards propagation), but the linear time
of each pass also makes the results applicable to forwards (inference)
performance. However, parallel linear recurrence can only accelerate inference
in scenarios where the entire input sequence is known at the start of the
inference phase.  We controlled for GPU memory usage within these experiments by
fixing $bT = 65,536$ for minibatch size $b$ and sequence length $T$, and chose a
popular architecture consisting of two stacked RNN layers with 256 hidden units
and an input size of 4.

Table \ref{table:rnn-throughput} shows that the
throughput advantage from using parallel linear recurrence compared to
serial linear recurrence reaches up to 9x. Simpler architectures
(for which the linear recurrence is a higher proportion of the
total computational load) are more affected by the switch to the parallel kernel.
This is
particularly clear in the case of the QRNN, where including wider
convolutional filters results in more time spent outside of the linear recurrence and
therefore reduces the speedup from linear recurrence parallelization.
\begin{table}[t]
  \caption{Parallel kernel speedup for a variety of LS-RNNs, implemented as two
    stacked RNN layers with 256 hidden units. We keep the GPU memory usage constant
  by fixing $bT = 65,536$ for minibatch size $b$ and sequence length $T$}
\begin{center}
\begin{tabular}{@{}lrrrr@{}}
\label{table:rnn-throughput}
Sequence Length & SRU & QRNN (filter size 2) & QRNN (filter size 10) & GILR-LSTM\\ \midrule
16 & 0.28 & 0.38 & 0.78 & 0.61\\
256 & 0.84 & 0.86 & 0.99 & 0.91\\
4,096 & 1.38 & 1.18 & 1.05 & 0.98\\
65,536 & 9.21 & 6.68 & 2.05 & 1.41\\ \bottomrule
\end{tabular}
\end{center}
\end{table}

% \subsubsection{GILR-LSTM throughput}

% \Eric{I think we should delete this section. Already over length limit, doesn't show
% anything great}

% Eric: I think we should delete this. We're over length, results aren't great.

% Finally, we compare the throughput of the GILR-LSTM to the current fastest
% LSTM implementation (which comes from the CuDNN library). Figure
% \ref{fig:tp_perf} shows that the GILR-LSTM has higher throughput for most of the
% small minibatch regime, with a 30x speedup at a sequence length of 65,536.
% The GILR-LSTM allows us to achieve high throughput for all choices of minibatch size and
% sequence length.

% Unlike the previous case, these speedups do come at a cost. The GILR-LSTM
% architecture is different to the canonical LSTM, and it's possible that the loss
% of model capacity outweighs the speedup in training. The following experiment
% shows that the GILR-LSTM is able to outperform the LSTM on an example originally
% designed to illustrate the advantages of the LSTM.

% \Eric{Chris: Put title on Figure \ref{fig:tp_perf}}

% \begin{figure}[t]
% \centering
% \includegraphics[width=12cm]{cudnn_heatmap.png}
% \caption{Throughput comparison between LSTM-256-256 with GILR-LSTM-256-256, with a
%   32-dimensional input and a 2-dimensional output, for various batch sizes and
%   sequence lengths.The LSTM only has a single row of data because its throughput
%   is independent of sequence length. Entries are missing from the GILR-LSTM table
%   because there was not enough memory on the GPU to handle such large batch
%   sizes and sequences.}
% \label{fig:tp_perf}
% \end{figure}

\subsection{Synthetic Experiment}
One of the key strengths of the LSTM is that it is capable of dealing with
long-term dependencies. In order to demonstrate that the GILR-LSTM is also able
to handle long-term dependencies we tackle a canonical example of inference over
many time steps from \citet{hochreiter1997long}.  We show that in fact the
GILR-LSTM is able to outperform the CuDNN-LSTM and extend to sequence lengths
orders of magnitude longer than dealt with previously. The input consists of
sequences of length $n$ where for $n > 0$ each element is a randomly chosen
one-hot vector \(x\) in $p$-dimensional space.  The first vector in each
sequence, \(x_0\), is always either \((1, 0, \ldots, 0)\) or \((-1, 0, \ldots,
0)\). The sequential model must read in an entire sequence and then output the
sign of the first sequence element.  This sequence classification problem
requires remembering the first element over the length of the sequence, and
early RNNs struggled with this for \(p\) as small as a few dozen. In the
original formulation of the problem (dealing in the regime with around one
hundred timesteps), the dimensionality of the input \(p\) is set equal to
\(n\). Since this would make the size of the input data grow impractically large
as \(\mathcal{O}(n^2)\) for long sequences, we fix \(p = 128\) as we vary \(n\).
\begin{figure}[] \centering
\includegraphics[width=12cm]{synthetic_diagram.pdf}
\caption{The structure of the synthetic example and the GILR-LSTM architecture
we used to tackle it. We feed in one-hot unit vectors \(x\) which are chosen
uniformly at random (with replacement). The class is determined by the very
first vector \(x_0\), which has a fixed direction. The sign of \(x_0\)
determines the class. In the diagram, each rounded block indicates a cell of the
RNN, whilst the square indicates a linear unit.}
\label{fig:synthetic_diagram}
\end{figure} We generated sequences for $n$ equal to 1,024, 8,192, and
1,048,576. For each of these we compared a two layer GILR-LSTM with 512 hidden
units to a two layer LSTM with 512 hidden units\footnote{For the longest
sequence length, the number of hidden units was decreased to 64 for both
architectures so that the net could fit in memory.} per layer implemented by
CuDNN.

We ran all experiments on a NVIDIA K80 GPU, with five runs per configuration
allowing us to find the average and standard deviation of the time and number of
iterations to convergence. We continually generated random sequences to serve
as input data.  A brief search over learning rate
and batch size was carried out to find the parameters which allow the network to
converge most rapidly for all runs. The criterion for convergence was five consecutive
minibatches giving 100\% accuracy. The learning curves
in figure \ref{fig:synthetic_training} give support to this being a reasonable
convergence criteria.
For the longest sequence length, we did not observe the CuDDN converging, even after
several days' training.
\begin{table}[]
\centering
\caption{Performance of the GILR-LSTM compared to the CuDNN-LSTM
  on problem 2b from \citet{hochreiter1997long}. We see that the GILR-LSTM has a
  clear advantage in training speed as the sequence length increases.  We did
  not observe the Cudnn-LSTM converging for the longest task, even after several
  days.}
\label{my-label}
\begin{tabular}{@{}lllllll@{}} \toprule Sequence Length &
\multicolumn{2}{c}{\textbf{1,024}} & \multicolumn{2}{c}{\textbf{8,192}} &
\multicolumn{2}{c}{\textbf{1,048,576}} \\ \midrule &CuDNN& GILR &
CuDNN &GILR&CuDNN&GILR\\ \cmidrule(l){2-7} Iterations (1000s) &
                     1.0 %bs=8
                     \(\pm\) 0.4 & 0.55 %bs=8
                                    \(\pm\) 0.04 & 0.44 %bs=8
                                                   \(\pm\) 0.05 & 0.56 %bs=8
                                                                 \(\pm\) 0.16
                                                               &-& 14 \(\pm\) 3 \\
\begin{tabular}[c]{@{}l@{}}Wall time (hours)\end{tabular} &
0.28 \(\pm\) 0.08 & 0.031 \(\pm\) 0.002 & 0.58 \(\pm\) 0.06 & 0.10 \(\pm\) 0.03
&-& 9.7 \(\pm\) 1.7 \\ \bottomrule
\end{tabular}
\end{table}

The results illustrate that the GILR-LSTM is able to converge between 6 and 10
times faster than the CuDNN-LSTM. This is somewhat surprising
given the LSTM was specifically constructed for problems of this sort, and the
CuDNN-LSTM is a highly-optimised implementation of the LSTM (to the extent that
the monolithic interface it exposes is difficult to understand or extend). The
GILR-LSTM is written almost entirely in native tensorflow, with only the call to
the linear recurrence kernel activating an outside hook. This GILR-LSTM
convergence leads to the conclusion that the non-linearities present in LSTM are
not necessary for solving this instance of the long-term dependency problem.
The time to convergence further leads to the conclusion that inclusion of a
non-linearity at every step incurs a significant training time slowdown.
Furthermore, the GILR-LSTM is able to learn to carry dependencies over a one
million element sequence. As far as we know, this one million step sequence
experiment is the longest sequential learning problem to be handled by neural
networks to date.
